---
title: "RProgAssingment3"
author: "Thomas J. Keller, PhD"
date: "August 27, 2014"
output: html_document
---

This is the documentation for August 2014 Coursera course 'RProgramming', assingment 3.

## Data source notes   
* Original data at http://hospitalcompare.hhs.gov 
* Data apparently from 12 months prior to July 2012  
* Subsets extracted by Coursera 'R Programming' staff, August 2014  
** https://d396qusza40orc.cloudfront.net/rprog%2Fdata%2FProgAssignment3-data.zip  
```{r}
dir("Assignment3-data/")
```

* The 'Hospital_Revised_Flatfiles.pdf' is the 'code book' for the csv data files.   
* outcome-of-care-measures.csv contains 30-ay mortality and readmission rates for  
** 'Heart.Attack', 'Heart.Failure' and 'Pneumonia'  
* "hospital-data.csv" contains information about the hospitals included in the study

### Background to data  
* including to determine whether hospitals should be fined for the quality of care  
* http://goo.gl/JAXFX  

## Focus of Assigment
* variables for 'Table 19' ("outcome-of-care-measures.csv")  
* variables for 'Table 11' ("hospital-data.csv")  

### Data exploration  
* number of columns  
* number of rows  
* sample the relevant data (columns: 11, 17, 23)  
```{r echo=FALSE}
outcome <- read.csv("Assignment3-data/outcome-of-care-measures.csv", colClasses="character")
ncol(outcome)
nrow(outcome)
colnames(outcome)
# the columns of interest are 2, 7, 11, 17 and 23 for respectively:
## Hospital.Name, State, Heart.Attack mortality, Heart.Failur mortality and Pneumonia mortality
outcome[1:20,c(2, 7, 11,17,23)]
```

#### Replace 'Not Available' values with NA
```{r}
## re-read the csv file, replacing 'Not Available' with NA
outcome <- read.csv("Assignment3-data/outcome-of-care-measures.csv", na.strings = "Not Available")
outcome[1:20,c(2, 7, 11,17,23)]
```

### subset the columns of interest, and rename columns for convenience  
```{r}
mortality <- outcome[, c(2, 7, 11,17,23)]
colnames(mortality) <- c("hospital", "state", "heart attack","heart failure","pneumonia")
str(mortality)
```


* histogram of '30-day death rates'  
```{r echo=FALSE}
hist(as.numeric(outcome[,11]), main="30-day Death Rate for All Hospitals")
hist(as.numeric(mortality$heart_attack))
```


## Finding the best hospital in a state
```{r}
best <- function(state, outcome) {
      ## read outcome data in downloaded csv file
      
      ## check that 'state' and 'outcome' arguments are valid; stop() and
      ### return "invalid state", or "invalid outcome" for bad input,
      ### "missing argment" if missing
      
      ## return hospital in 'state' with lowest 30-day death rate (mortality)
      ### example best("TX", "heart attack") returns [1] "CYPRESS FAIRBANKS..."
      ### hospital names should be sorted alphabetically, return the first for ties
      
}
```

## Ranking hospitals by outcome in a state  
```{r}
rankhospital <- function(state, outcome, num = "best") {
      ## read outcome data in downloaded csv file
      
      ## check that 'state' and 'outcome' arguments are valid; stop() and
      ### return "invalid state", or "invalid outcome" for bad input,
      ### "missing argment" if missing argument)
      ### "best" is rank 1, "worst" is the largest number for rank for the state
     
      
      ## return hospital in 'state' with that rank for 30-day death rate (mortality)
      ### a num (rank) that is greater than the largest number for rank should return NA
      
}
```

## Ranking hospitals in all states
```{r}
rankall <- function(outcome, num = "best") {
      ## read outcome data in downloaded csv file
      
      ## check the 'outcome' arguments is valid; stop() and
      ### return "invalid outcome" for bad input,
      ### "missing argment" if missing argument)
      ### "best" is rank 1, "worst" is the largest number for rank for the state
      
      ## for each state, find the hospital of the given rank
      ### NA if no hospital for a state has that rank
      
      ## return a data.frame with 2 columns: the hospital names and state (Abreciation)
      ### exclude states with NA for the rank
      
      
}
```

### functional units
* read outcome and subset relevant columns, replace 'Not Available' with NA
* check arguments
** 'state' must be a member of the mortality$state list of unique values
** check 'outcome' arg is c("heart attack", "heart failure", "pneumonia")  
```{r}
states <- as.character(unique(mortality$state))
diseases <- c("heart attack", "heart failure", "pneumonia")
check <- function(state, outcome) {
 if(missing(state)) stop("missing argument")
 if(missing(outcome)) stop("missing argument")
 ## allow variations in case
 state = toupper(state)
 outcome = tolower(outcome)
 if(is.na(states[match(state,states)])) stop("invalid state")
 if(is.na(diseases[match(outcome,diseases)])) stop("invalid outcome")
 args <- paste(outcome,state, sep= " in ")
 return(args)
}

```

* subset by state
* best outcome  
** be sure to include validation in the final function  
```{r}
stateoutcome <- function(state, outcome) {
      ## fix outcome names
      disease <- sub("\\s+","_",outcome)          
      ## fix column names
      colnames(mortality) <- c("hospital", "state", "heart_attack","heart_failure","pneumonia")
      mystate.df <- mortality[mortality$state == state,]
      stateorder <- order(mystate.df[,which(names(mystate.df) == disease)],mystate.df[,1])
      besthospital <- as.character(mystate.df[stateorder[1],1])
      return(besthospital)
}
```

* subset by state
* rank the outcome  
** be sure to include validation in the final function  
```{r}
rankhospital <- function(state, outcome, num = "best"){
      ## validate arguments
       if(missing(num)) stop("missing argument")
      ## fix outcome names
      disease <- sub("\\s+","_",outcome)   
      ## fix column names
      colnames(mortality) <- c("hospital", "state", "heart_attack","heart_failure","pneumonia")
      mystate.df <- mortality[mortality$state == state, c(1,which(names(mortality) == disease))]
      stateorder <- order(mystate.df[,which(names(mystate.df) == disease)],mystate.df[,1], na.last = T)
      ## recode "best" to 1 and "wworst" to length(stateorder)
      num <- ifelse(num == "best", 1, num)
      num <- ifelse(num == "worst", length(stateorder), num)
      num <- as.numeric(num)
      if(is.na(num)) stop("invalid num")
      ## order by outcome and hospital name
      mystate.df <- mystate.df[stateorder[num],c(1,which(names(mystate.df) == disease))]
      return(mystate.df)
}
```

* apply to all states
* rank outcome and order by outcome and hospital name
** return hospital name and state
```{r}
rankall <- function(outcome, num = "best") {
      states <- as.character(unique(mortality$state))
      
      myfun <- function(state, outcome, num = "best"){
      ## validate arguments
       if(missing(num)) stop("missing argument")
      ## fix outcome names
      disease <- sub("\\s+","_",outcome)   
      ## fix column names
      colnames(mortality) <- c("hospital", "state", "heart_attack","heart_failure","pneumonia")
      mystate.df <- mortality[mortality$state == state, c(1,which(names(mortality) == disease))]
      stateorder <- order(mystate.df[,which(names(mystate.df) == disease)],mystate.df[,1], na.last = T)
      ## recode "best" to 1 and "wworst" to length(stateorder)
      num <- ifelse(num == "best", 1, num)
      num <- ifelse(num == "worst", length(stateorder), num)
      num <- as.numeric(num)
      if(is.na(num)) stop("invalid num")
      ## order by outcome and hospital name
      mystate.df <- mystate.df[stateorder[num],c(1,which(names(mystate.df) == disease))]
      }
      rowbind(output) <- lapply(seq_along(states),myfun)
      return(output)
}
```

